---
title: "Assignment 2A DATA 607"
format: html
editor: visual
---

## Approach 2A

For this assignment, I plan to collect 1–5 ratings for six movies from at least five people and store them in a SQLite database using separate tables for users, movies, and ratings. I will then load the data into R and explore online resources and LLM tools to determine appropriate methods for analyzing and summarizing the ratings. Since not everyone will have seen every movie, I will treat those ratings as missing values (`NA`) rather than assigning them a score.

Install Postgres packages so I can use the sql data

```{r}
library(DBI)
library(RPostgres)
```

```{r}
packageVersion("DBI")
packageVersion("RPostgres")
```

```{r}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "movie_ratings",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = rstudioapi::askForPassword("PostgreSQL password")
)
```

```{r}
movie_ratings <- read.csv("movie_ratings_2A.csv")

head(movie_ratings)
```

```{r}
str(movie_ratings)
summary(movie_ratings)

aggregate(rating ~ movie_title,
          data = movie_ratings,
          FUN = mean)
```

```{r}
rating_matrix <- xtabs(
  rating ~ user_name + movie_title,
  data = movie_ratings,
  sparse = FALSE
)

rating_matrix <- reshape(
  movie_ratings,
  idvar = "user_name",
  timevar = "movie_title",
  direction = "wide"
)

rating_matrix
```
